Can We Crunch the Social Graph on the Cheap?

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
almost all of eval condensed

%% move stuff to defs file
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
%% \renewcommand{\todo}[1]{\relax}

%\documentclass{acm_proc_article-sp}
\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage{comment}
\begin{document}

@@

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}

\maketitle
\begin{abstract}
% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. These irregular
% problems present a challenge for parallel machines: obtaining
% performance is not straightforward. Multithreading works well for
% such problems, but the only large-scale implementation of the technique
% is the Cray XMT, an expensive, non-commodity machine.

% Our goal is to build a system with XMT-like properties, implemented
% with commodity components. This paper presents a runtime for latency
% tolerance, along with our plans for scaling up to multiple
% nodes. Our analysis shows the feasibility of our approach.

  Crunching large graphs is the basis of many emerging applications,
  such as social network analysis and bioinformatics. Graph analytics
  algorithms exhibit little locality and therefore present significant
  performance challenges. Hardware multithreading systems (e.g, Cray
  XMT) showed that with enough concurrency, we can tolerate long
  latencies. Unfortunately, this solution is not available with
  commodity parts.

  Our goal is to develop a latency-tolerant system built out of
  commodity parts and mostly in software. The proposed system includes
  a runtime that supports a large number of lightweight contexts,
  full-empty synchronization and a memory manager that provides a
  high-latency but high-bandwidth global shared memory. This paper
  lays out the vision for our system, and justifies its feasibility
  with a performance analysis of the runtime for latency tolerance.

% Our analysis shows the feasibility of our
%   approach.

% To that end we developed a new runtime system that supports
%   a large number of lightweight contexts, together with full-empty
%   synchronization primitives and a large global address space.

% To that end, we propose a
%   new runtime system that is able to sustain a large number of active
%   threads

%  XMT-like properties, implemented
%   with commodity components. This paper presents a runtime for latency
%   tolerance, along with our plans for scaling up to multiple
%   nodes. Our analysis shows the feasibility of our approach.

%   Large, sparse, irregular graphs map poorly to the memory systems of
%   commodity hardware. The scale of emerging real world datasets
%   presents a challenge for straightforward algorithms which will run
%   inefficiently due to many high-latency accesses. Pervasive
%   parallelism can tolerate that latency, but the only large-scale
%   implementation of the technique is the Cray XMT, an expensive,
%   non-commodity machine.

%   If we can adapt the XMT's programming model to commodity hardware,
%   we can dramatically improve performance without resorting to
%   expensive custom systems. This paper presents a runtime for latency
%   tolerance through the use of parallel lightweight contexts, along
%   with our plans for scaling up to multiple nodes. Our analysis shows
%   the feasibility of our approach.
\end{abstract}

@@

Many important emerging applications such as social network analysis,
bioinformatics and analysing vast sensor data from actual networks
(e.g, water, electrical grid) are based on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. For example, in January 2011, Facebook had
500 million active users with an average of 130 friends each
\cite{Facebook:2011p91. Its friend suggestion algorithm in 2010 ran on
  a 40-node cluster with 72GB memory per node, and still, new
  suggestions could only be computed every two days
  \cite{Backstrom:2010p90}. Speeding up these applications at a low
  cost would have a large impact in how we deal and react to the fast
  growth of the data-set and would make the results even more
  valuable.

% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. \todo{more!}

@@

In this context, the most interesting computational challenge comes
from large, low-diameter, power law graphs: this combination makes
extracting performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near
every other vertex. Their power law distribution means that a few
vertices generate much more work than the rest, leading to load
imbalances. Together, these properties lead to a difficult
conclusion: each edge traversal is likely to lead to expensive
inter-node communication. %cache miss or remote access.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem (or
communication problem) through concurrency rather than caching. Each
XMT processor supports 128 hardware contexts and 1024 outstanding
memory operations, and is able to switch contexts every cycle.
% \todo{cite XMT description}.  %% LUIS: there is enough information here.
This ability comes at a cost: the XMT is an expensive, non-commodity
machine. Moreover, arguably a lot of XMT's advantages for these type
of applications come at the cost of performance in general-purpose
applications.

% \todo{don't we want to also say XMT has poor performance on x
%   kind of apps?}

@@

%We want the best of both worlds. 
We believe we can build a system mostly based on commodity parts that
can attain XMT-like performance with a familiar, also XMT-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach also has the added
benefit of not sacrificing general-purpose performance.

Figure~\ref{fig:system-overview} shows the structure of our
proposal. It is composed of multiple nodes built with commodity
processors, communicating over an Infiniband network. We add two
components: a runtime system, responsible for executing the user's
threads, and a global memory manager, responsible for facilitating
memory requests to the non-coherent global memory shared across the
nodes.  We are exploring a mix of hardware (FPGA) and software
implementations of the memory manager.

This paper describes the vision of our system and explores the
feasibility of our ideas using a single-node implementation of a
lightweight threading runtime. We use prefetch instructions together
with lightweight threads to tolerate the latency of DRAM access on the
local node node. We use adversarial pointer chasing benchmarks to
verify our runtime's overhead is acceptable. We also model the effects
of pointer chasing on a remote node by artificially increasing
latencys to the order of several micro-seconds.

The rest of the paper is organized as follows. We quickly describe our
programming model in section~\ref{sec:model}. We describe the
implementation of our latency-tolerant runtime and plans for extending
to multiple nodes in section~\ref{sec:approach}. We evaluate our
runtime in section~\ref{sec:evaluation}.  We present related work in
section~\ref{sec:related} and conclude in
section~\ref{sec:conclusion}.

@@

Our goal is to preserve the XMT programming model: large shared global
address space, explicit concurrency with a large number of threads,
and full-empty-bits synchronization.  We partition the overall address
space into a global shared address space and per node private address
spaces. Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node. In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

% Whereas in the PGAS model coherency of global copies is implictly
% maintained across the system, % \todo{cite pgas?}
% we completely forego
% caching of globally shared data.

A node can exploit temporal locality of data stored in the global
space only by first copying it into the node's local space.  In this
way, a node can access data at random across the system while
incurring no hardware coherence overhead: consistency is enforced by
our mechanisms for explicit global synchronization.

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerant on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g. a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation.  Concurrency
is also used -- via aggregation of requests to a common node
destination -- to make more efficient use of network bandwidth and in
some cases reduce bandwidth demands.

Synchronization on locations in the node private address space is the
same as on conventional systems.  Synchronization on global addresses
works differently: we provide atomic operations such as {\tt
  int\_fetch\_add} as well as operations using full-empty-bits, as on
the Cray XMT.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield: atomic operations executed as
x86 instructions sabbotage concurrency by forcing all outstanding
memory references to complete before new ones can be issued.  By
yielding, the core can switch to another computation while the
synchronization is accomplished outside of the pipeline.  In addition,
the implementation can exploit aggregation when concurrency is
abundant to increase the throughput of global synchronization.

%% I dont think we need the paragraph below.

% Thus, the programming model provides the programmer with options: express
% locality within the private space, and it will be exploited to reduce
% latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
% tolerating latency and reducing the cost of bandwidth.

@@

Figure~\ref{fig:multi-green} shows the result. For this
experiment, we limited memory concurrency to one concurrent miss per
coroutine, so the only source of memory concurrency is the use of
coroutines. We are able to obtain a rate of 275 million references per
second with 48 concurrent misses, or 8 coroutines per core. These
results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36. When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. We
also observe some points that do not follow the trend and look like noise, but these results are actually repeatable. This effect is harder to explain, but we suspect it is
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

% cache pressure figure
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.42\textwidth]{figures/cache-pressure-edited.pdf}
  \end{center}
  \caption{Cache pressure with coroutines.}
  \label{fig:cache-pressure}
\end{figure}

We ran this code on six cores with all data allocated in the same
processor's memory. We varied the size of the local working set from 2KB
to 4MB, and ran both 0 and 32 local updates for each pointer
traversal. We measured the overall reference rate, including both the local updates and remote references.

Figure~\ref{fig:cache-pressure} shows the result. On the left of the
figure, we see the list-traversal-only case, where we allocate
per-coroutine space but make no references to it. As we would expect,
the reference rates grow with the number of coroutines just as in
Figure~\ref{fig:multi-green}. On the right of the figure, we see the
case with 32 updates for each list traversal. While performance
decreases for large working set sizes and large number of coroutines,
we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
core. \todo{preempt questions about drop at 16 which is much less than the about 50 needed per core by saying that the drop is likely the same drop effect seen earlier, which is likely to be prefetch squashing since it happens at number of line fill buffers}

\subsection{Multi-node}

To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency. First we measure the QPI bandwidth between the two chips of the test machine, since the chip would talk to the network interface over a QPI link.

% take this figure out if we need the space
% the main point is room for remote references to local memory
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
  \end{center}
  \caption{Pointer chasing in a remote processor's memory.}
  \label{fig:listwalk-qpi}
\end{figure}
%%%%%%%%%%

We measured the rate at which the cores can make requests
over the QPI link. For this experiment, we allocated the lists in the
second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.

>>>>

%% move stuff to defs file
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
 \renewcommand{\todo}[1]{\relax}

\newcommand{\mrps}[0]{Mref/s}

%\documentclass{acm_proc_article-sp}
\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage{comment}
\usepackage[labelfont=bf,font=small,belowskip=-7pt,aboveskip=-10pt]{caption}
\begin{document}

@@

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}

\maketitle
\begin{abstract}

 Techniques from graph analytics have found important applications in
  areas such as social networks and bioinformatics. These irregular
  problems present a challenge for parallel machines: obtaining
  performance is not straightforward. Multithreading works well for
  such problems, but the only large-scale implementation of the technique
  is the Cray XMT, an expensive, non-commodity machine.

  Our goal is to build a system with XMT-like properties, implemented
  with commodity components. This paper presents a runtime for latency
  tolerance, along with our plans for scaling up to multiple
  nodes. Our analysis shows the feasibility of our approach.
\end{abstract}

@@

Techniques from graph analytics have found important applications in
areas such as social networks and bioinformatics. \todo{more!}

@@

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near every
other vertex. Their power law distribution means that a few vertices
generate much more work than the rest, leading to load
imbalances. Together, these properties lead to a
difficult conclusion: each edge traversal is likely to induce a very
expensive cache miss or remote access.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}

@@

We want the best of both worlds. Our goal is to build a system that
has good performance on these low-locality graph codes but is
implemented with cheap commodity processors.

Figure~\ref{fig:system-overview} shows the structure of our
proposal. It is composed of multiple nodes built with commodity
processors, communicating over an Infiniband network. We add two
components: a runtime system, responsible for executing the user's
threads, and a global memory manager, responsible for facilitating
memory requests to the non-coherent global memory shared across the
nodes.  We are exploring a mix of hardware and software
implementations of the memory manager.

This paper explores the feasibility of our multi-node goal using a
single-node implementation of our lightweight threading runtime. We
use prefetch instructions together with lightweight threads to
tolerate the latency of DRAM access on the local node node. We use
pointer chasing benchmarks to verify our runtime's overhead is
acceptable. We also model the effects of pointer chasing on a remote
node.

The rest of the paper is organized as follows. We describe our
proposal from a programmer's viewpoint in section~\ref{sec:model}. We
describe the implementation of our latency-tolerant runtime and plans
for extending to multiple nodes in section~\ref{sec:approach}. We
evaluate our runtime in section~\ref{sec:evaluation}.  We present
related work in section~\ref{sec:related} and conclude in
section~\ref{sec:conclusion}.

@@

Our shared-memory programming model partitions the overall address
space into a global shared address space and per node private address
spaces.  Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node.  In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

Whereas in the PGAS model coherency of global copies is implictly
maintained across the system \todo{cite pgas?}, we forego caching of globally shared
data.  A node can exploit temporal locality of data stored in
the global space only by first copying it into the node's local space.
In this way, a node can access data at random across the system while
incurring no hardware coherence overhead: consistency is enforced by
our mechanisms for explicit global synchronization.

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent.  Our programming model serves to
augment threaded programming languages so that they become latency
tolerant on large address spaces.  Software multithreading is used to
provide latency tolerance: a computation that is about to execute a
long latency operation such as a remote memory reference initiates the
operation and then yields to the scheduler, which quickly resumes
execution of another computation.  Concurrency is also used -- via
aggregation of requests to a common node destination -- to make more
efficient use of network bandwidth and in some cases reduce bandwidth
demands.

Synchronization on locations in the node private address space is the
same as on conventional systems.  Synchronization on global addresses
works differently: we provide atomic operations such as {\tt
  int\_fetch\_add} as well as operations using full-bits, as on the
Cray XMT.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield: atomic operations executed as
x86 instructions sabbotage concurrency by forcing all outstanding
memory references to complete before new ones can be issued.  By
yielding, the core can switch to another computation while the
synchronization is accomplished outside of the pipeline.  In addition,
the implementation can exploit aggregation when concurrency is
abundant to increase the throughput of global synchronization.

Thus, the programming model provides the programmer with options: express
locality within the private space, and it will be exploited to reduce
latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
tolerating latency and reducing the cost of bandwidth.

@@

Figure~\ref{fig:multi-green} shows the result. For this
experiment, we limited memory concurrency to one concurrent miss per
coroutine, so the only source of memory concurrency is the use of
coroutines. We are able to obtain a rate of 275 \mrps with 48 concurrent misses, or 8 coroutines per core. These
results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36. When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions. We
also observe many points that do not follow the trend are below the max, but these results are repeatable. This effect is harder to explain, but we suspect it is
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

% cache pressure figure
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.42\textwidth]{figures/cache-pressure-edited.pdf}
  \end{center}
  \caption{Cache pressure with coroutines, on six cores.}
  \label{fig:cache-pressure}
\end{figure}

We varied the size of the local working set from 2KB
to 4MB, and ran both 0 and 32 local updates for each pointer
traversal. We measured the overall reference rate, including both the local updates and remote references.

Figure~\ref{fig:cache-pressure} shows the result. As we would expect,
when there are no references to the local structures, the reference rates grow with the number of coroutines just as in
Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance
decreases for large working set sizes and large numbers of coroutines,
we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
core. \todo{preempt questions about drop at 16 which is much less than the about 50 needed per core by saying that the drop is likely the same drop effect seen earlier, which is likely to be prefetch squashing since it happens at number of line fill buffers}

\subsection{Performance with multiple nodes}

To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency. Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.

% take this figure out if we need the space
% maybe even take out whole QPI thing and just relate the 100M?
% the main point is room for remote references to local memory
%\begin{figure}[h]
 % \begin{center}
  %  \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
 % \end{center}
 % \caption{Pointer chasing in a remote processor's memory.}
 % \label{fig:listwalk-qpi}
%\end{figure}
%%%%%%%%%%

For this experiment, we allocated the lists in the
second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.
